---
title: "PSHB Models"
output: html_document
date: '`r format(Sys.Date(), "%d/%m/%Y")`'
bibliography: 'PSHBRefs.bib'
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo=TRUE)
options(scipen = 10, digits = 2)
```

## Basic life history

Polyphagous shothole borer, *Euwallacea fornicatus*, is a pest of a wide range of plants and has recently expanded its range from south-east Asian to Africa, the Americas, and Australia.

Mated females alight on a host plant and bore a small gallery in the host tree tissue. They seed an ambrosia fungus into this gallery and spend some amount of time building the gallery before laying eggs. After hatching, juveniles moult through several larval instars, pupate, and eventually eclose as adults. The sex ratio of adults is m:f 1:3-5. Males are wingless and never leave the gallery. Females are winged. They mate with their brothers, and then exit the gallery to found a new gallery (on the same or different host). The rate of egg laying declines steadily over time, with females eventually dying in the gallery rather than forming a new one.

Development of all stages is strongly temperature dependent. There is some evidence of developmental deaths occurring between hatching and pupation. Other sources of mortality are likely at the dispersal phase (predation, misadventure), otherwise the life history is fairly safe from predation.

## Our treatment

### Discrete time model

The life history suggests we might profitably partition the population of females as:

  * $J(t,x)$, the juvenile stages (eggs through pupae): some mortality in the nest, no dispersal 
  * $P(t,x)$, the pre-adult stage (the dispersal phase): dispersal, with some mortality 
  * $A(t,x)$, the adult stage (breeding females): no dispersal, mortality through senescence.

We measure time, $t$, in days. We also partition our population among host trees, $x$.

#### Within host dynamics

Within a single host (i.e.ignoring $x$), and using the same state variables above we would have:


\begin{align}
J(t+1) &= \phi_J(1-\alpha_J)J(t) + fA(t\\
P(t+1) &= \phi_J\alpha_JJ(t) + \phi_P(1-\alpha_P)(1-\mu)P(t) + 0 \\
A(t+1) &= 0 + \phi_P\alpha_P(1-\mu)P(t) + \phi_AA(t)
\end{align}

Where:

  * $f(t)$ is the mean fecundity of a female in the population in an interval of time 
  * $\alpha_i(t)$ are transition probabilities, describing transition between states
  * $\phi_i(t)$ are survival probabilities, describing probability of surviving an interval of time in each state 
  * $\mu(t)$ is the probability of dispersing away from the host tree.

Although we drop the notation, all these are functions of time, to capture the effect of temperature on the vital rates.  In matrix form, we first define the column vector $\mathbf{n}= \{J, P, A\}$. Then we have:

$$
\begin{gather}
 \mathbf{n}(t+1) =
  \begin{bmatrix}
   \phi_J(1-\alpha_J) & 0 & f \\
   \phi_J\alpha_J & \phi_P(1-\alpha_P)(1-\mu) & 0 \\
   0 & \phi_P\alpha_P(1-\mu) & \phi_A
  \end{bmatrix}
 \mathbf{n}(t)
\end{gather}
$$

The matrix shown is the within-population transition matrix, $\mathbf{W}$.

Note also that in the within-host model, we currently neglect to account for the arrival of pre-adults from other hosts. We will come back to this in due course.

Note that there is no density dependence in the model for now. The obvious place to put this would be on $\phi_J$. That is, we could make $\phi_J$ some negative function of total number of juveniles produced in the host. A solution here is a simple threshold such that,

$$
\begin{gather}
  \phi_J(t, x) = 
  \begin{cases}
    \phi_J & \text{if } \sum_t{fA(t, x)} < C(x) \\
    0 & \text{if } \sum_t{fA(t, x)} > C(x)
  \end{cases}
\end{gather}
$$ 

Here, each host could have its own threshold, $C(x)$, for production of beetles. Beyond this threshold, the host resource is used up and $\phi_J$ collapses to 0. This threshold will also be useful later when we describe dispersal dynamics.

#### Between host dynamics

We now seek to extend the model to space, to capture dynamics across multiple hosts. To do this, we need to describe dispersal dynamics. We assume that $\mu(t, x) P(t, x)$ leave each host at each time period. We might further assume that there is some survival rate associated with dispersing between hosts and establishing a new gallery. Let's call that $\phi_\mu$. Thus, the number of effective dispersers from a given host at a given time is: $\phi_\mu\mu(t,x) P(t, x)$.

We will make this model spatially explicit. So the probability of arriving at a given host will depend on the distance between source and recipient host, let us call that distance, $d_{xy}$, the distance between host $x$ and host $y$ (where $y \neq x$). We can imagine a pairwise distance matrix, $\mathbf{D}$, which is an $h\times h$ matrix, where $h$ is the total number of hosts. 

We now need to work out the probability of moving to a given host.  We will assume a gravity model for dispersal [@jongejans2015], in which the attractiveness of hosts as a dispersal destination varies over time and space. We first define the 'gravitational mass' of a host as the probability that pre-adults will not disperse away from that host:

$$
g(t, x) = 1-\mu(t, x) 
$$
We make $g(t, x)$ density dependent in the following way:

$$
g(t, x) = (1-\delta e^{-\zeta A(t, x)})s(t, x),
$$
where $\delta$ is a parameter setting the minimum probability of dispersing away from the host (which we may make host specific later), $\zeta$ is another parameter defining the strength of density dependence on dispersal, and $s(t, x)$ is an indicator variable that collapses to $g(t,x) = 0$ when the host threshold is reached.  That is,

$$
s(t, x) = 
  \begin{cases}
    1 & \text{if } \sum_t{fA(t, x)} < C(x) \\
    0 & \text{if } \sum_t{fA(t, x)} > C(x)
  \end{cases}
$$

This function, $g(t, x)$ adds an additional two parameters.  It varies over time and host and gives us a host gravity that increases with increasing density of adults, and then collapses to zero when the host has been fully exploited.

For each time step, we can calculate the gravitational mass of each host.  For infected hosts, we need to work out where the dispersers go to.  We define the probability of arrival at host $y$ from host $x$ as:

$$
a(y, x, t) = k(t,x)\frac{g(t, y)}{d_{xy}^2}.
$$

This is a gravitational model in which each host has a gravitational mass that affects how attractive it is to dispersing beetles.  The influence of this attractional force decays with the square of distance. $k(t,x)$ is a normalisation function for each host at each time step:

$$
k(t,x) = (1-g(t,x))\sum_y{\frac{d^2_{xy}}{g(t,y)}}
$$

So, at each time step, we calculate the $h$-length vector, $\mathbf{g}$, we can then calculate the vector $\mathbf{k}$, we then use these to build the $h \times h$ matrix $\mathbf{A}$, which holds all the probabilities of moving from $x$ to $y$.  Note that the diagonal of $\mathbf{A}$ is given by the vector, $1-g(t,x)$.


It is tempting at this stage to specify dispersal as a stochastic process. The reason for this is that without stochasticity we can have fractions of individuals arriving at sites and growing; we would need to control this in some way. With stochasticity, we can arrange things to only have integer values of immigrants. We can draw the realised number of dispersers coming from $x$ and arriving to each site, $y$ as,

$$
\mathbf{m_x} \sim \text{Multinom}(P(t, x), \phi_\mu a(t, x, y))
$$ where $m_y$ is the realised number of migrants emerging from site $y$:


We can get away with this without further adjustments to the within-host matrix, because we assume that it is survival during dispersal that introduces the stochastic element here.

By stepping through all sites, $s$, we arrive at a realised dispersal matrix, $\mathbf{M}$. And the realised number of immigrants arriving at site $y$ is then the vector given by the column sums of $\mathbf{M}$: $\mathbf{m_+} = \sum_x \mathbf m_x$.

Note, that Nick points out that using an alternative stochastic process (e.g. Poisson) will likely run a lot faster if this model is to be used for inference.

### Within and between host model

So, now we need to bring this all together. First step is to re-define the population vector as $\mathbf{n} = \{\mathbf{n_1}, ..., \mathbf{n_h} \} = \{J_1, P_1, A_1, J_2, P_2, A_2, ..., J_h, P_h, A_h,\}$, thus now containing all individuals at all stages across all the hosts in the population.

We then define the $h \times h$ diagonal block matrix, $\mathbf{T}$:

$$
\mathbf{T} = 
 \begin{bmatrix}
  \mathbf{W} & 0  & 0 \\
  0 & \ddots & 0 \\
  0 & 0 & \mathbf{W}
 \end{bmatrix}.
$$ 

Here each $0$ denotes a $3\times 3$ matrix of zeroes, and $\mathbf{W}$ is the within-host transition matrix.

Normally, we would place dispersal into the off-diagonal elements of this matrix, but we have made dispersal stochastic, so easier to leave it out. Instead, we then need to get $\mathbf m_+$ into the same length as our redefined $\mathbf{n}$. So we define a new vector,

$$
\mathbf{m} = \{0, 0, \mathbf{m_{+1}}, 0, 0, \mathbf{m_{+2}}, ..., 0, 0, \mathbf{m_{+h}} \}
$$ Our final model then becomes:

$$
\begin{gather}
 \mathbf{n}(t+1) = \mathbf{T}\mathbf{n}(t) + \mathbf{m}(t).
\end{gather}
$$

## Estimating life-history parameters

### Transition probabilities

```{r parameterSet, echo=FALSE}
delta_t <- 20
wal_juv <- 266
wal_pre <- 136
```


Several sources have estimated the 'thermal constant', $K$ for PSHB and related species.  $K$ is the cumulative number of degree days above some minimum temperature that it takes for development to complete.  It is a standard approach in entomology. So, if the minimum temperature is 15 degrees and we have a day where the mean temperature is 20 degrees, then we have accumulated 5 degree days towards development.  This is what is meant by $K$.

The results for $K$ in *Euwallacea* are nicely summarised in @dodge2019.

| Species | $T_{min}$ | $T_{opt}$ | $T_{max}$ | $K$ | Reference |
|:--:|:--:|:--:|:--:|:--:|:--:|
| TSHB | 15 | 30 | 32 | 373 | @walgama2007 |
| PSHB | 13.34 | 27.51 | 33.08 | 398 | @umeda2019 |
| KSHB | 12.77 | 28.04 | 31.99 | 318 | @dodge2019 |

Note Dodge's estimate is for (nearly) the full life cycle from egg to emergent adult (*i.e.*, $\alpha_J\alpha_P$). Umeda's estimate is for the transition from juvenile to pre-adult (*i.e.*, $\alpha_J$).  Walgama's estimate is for the full life-cycle.  Dodge and Walgama's studies are best, and Walgama's gives the clearest breakdown of rates.

This degree.day format is not actually useful to us, because while development can be seen as constant with respect to degree days, survival has only ever been estimated as a function of temperature.  My initial thought was to change the time axis of our model to degree.days and use the resulting constant transition probabilities.  This would have been computationally most efficient. Problem with this is that the degree.day axis obscures absolute temperature, and survival has only ever been estimated with regard to absolute temperature.  So if we want a model with both transition and survival probabilities, we need to stick to the absolute time axis.  

Thus, we need to step back from the degree day results, and look for the per day rates that went into the estimation of $K$.  @umeda2019 provide the following data in their paper.

```{r umedaData, fig.cap="Data on the rate of development from egg to pre-adult, from @umeda2019. Line is fitted ignoring the result for 32 degrees, which is considered above the thermal threshold."}
temp <- c(18, 20, 25, 30, 32)
rate <- c(0.013, 0.020, 0.0258, 0.0421, 0.0206)
plot(rate~temp, xlab = "Temperature", ylab = "Development rate (per day)")
ss <- 1:4
mod <- lm(rate[ss]~temp[ss])
abline(reg=mod)
```

We can use this to build a function returning $\alpha_J$. First, the estimated coefficients of the linear model:


```{r}
coef(mod)
```

Then we can build a function.

```{r alpha_J_temp}
alpha_J_temp <- function(temperature, lower = 13.5, upper = 31){
  rate <- -0.0273 + 0.0023*temperature
  prob <- 1-exp(-rate)
  ifelse(temperature>lower & temperature < upper, prob, 0)
}
```

Let's see what $\alpha_J$ looks like across a range of temperatures.

```{r alpha_J_plot}
temps <- seq(10, 35, 0.1)
alpha_J <- alpha_J_temp(temps)
plot(alpha_J~temps, 
     type = "l", 
     xlab = "Temperature", 
     ylab = quote(italic(alpha[J])),
     bty = "l")
```

@umeda2019 do not provide estimates of transition rates through the pre-adult stage.  Instead, here we turn to @walgama2007, who estimate 136 degree days with a minimum of 15 degrees for this stage.  These values can be mapped back to a linear function of development rate against temperature, as follows.

```{r alpha_P_estimate}
alpha_P_temp <- function(temperature, lower = 15, upper = 31){
  DD <- 136 #degree days from walgama
  beta <- 1/DD # implied slope of rate on temperature
  alpha <- -15*beta # implied intercept
  rate <- alpha + beta*temperature
  prob <- 1-exp(-rate)
  ifelse(temperature>lower & temperature < upper, prob, 0)
}

alpha_P <- alpha_P_temp(temps)
plot(alpha_P~temps, 
     type = "l", 
     xlab = "Temperature", 
     ylab = quote(italic(alpha[P])),
     bty = "l")

```


### Survival probabilities

Survival probability is also clearly temperature dependent.  @umeda2019 do not record survival, so again we are back to @walgama2007, who did.  Moreover, @walgama2007 are the only set of researchers who had nearly 100% survival at optimal temperature, indicating that their husbandry setup worked well.  Their data on survival are reported in their Figure 2.  I used WebPlotDigitizer to extract these data, and added on development time at each temperature.  This allows me to express mortality probability per day and examine how that changes with temperature.

```{r walgamaSurvival}
sDat <- read.csv(file = "../dat/walgamaFig2DataExtract.csv")
sDat$Mortality <- sDat$Mortality/100 # convert to probability of mortality over n days
sDat$Mortality[sDat$Mortality>1] <- 1 # catch the 1.002s
mortRate <- -log(1-sDat$Mortality)/sDat$Days # convert to a rate
mortProb <- 1-exp(-mortRate) # convert to a daily probability
plot(mortProb~sDat$Temperature)
```

This is a little awkward.  Let's express it as a survival rate instead, and see if we can fit a thermal performance curve of some kind.  We have plenty of data in this case, so can fit a decent thermal performance curve.  (I also happen to have a bunch of functions handy that fit TPCs.)

```{r survTPC}
survProb <- 1-mortProb
sDat <- cbind(sDat, mortProb, survProb)
plot(survProb~sDat$Temperature)

source("../src/TPCFunctions.R")
TPCmatrix <- cbind(sDat$Temperature, survProb)
TPCmatrix <- na.omit(TPCmatrix)
pars <- TPC.q.fit(TPCmatrix, in.acc = 0.3)$par
survProbPred <- TPC.q(temps, rmax = pars[1], Trmax = pars[2], acc = pars[3], dec.prop = pars[4])
lines(survProbPred~temps)
```

So we now have a function for juvenile survival rate. This can be applied to the juvenile stage.

```{r}
phi_J <- function(temperature){
  TPC.q(temps, rmax = 0.99, Trmax = 29.5, acc = 80, dec.prop = 0.15)
}
```

Pre-adult and adult survival appears to be less affected by temperature.  Let us assume that this is constant.  @gadd1941 estimates a maximum laying lifespan for a female as 49 days, with a rough estimate of the mean laying lifespan of 32 days. This would yield a per day mortality rate of 1/32.  This equates to a per day survival probability of $\phi_P = \phi_A=$ `r exp(-1/32)`.

We can assume pre-adult survival is the same as adult survival. This leaves us only with the mortality cost associated with dispersal, $\phi_\mu$, which we have no measure for.

### Fecundity

@gadd1941 estimates the rate of egg laying per female at 0.927 eggs per day per female.  His data also has a sex ratio of m:f = 81:237.  So the fraction of eggs that are female is `r 237/(237+81)`.  Hence, the rate of egg laying of *female* eggs is likely to be around `r 0.927*237/(237+81)`.

## Estimating priors

All this is useful for point estimates for a deterministic model, but we are also interested in a Bayesian inference model.  To start on that, we will need decent priors for all the above. 

### Temperature dependent vital rates

We propose that an efficient strategy here will be to estimate a family of temperature performance curves (TPCs).  With a curve for each of the temperature-dependent parameters: $\alpha_J(T)$, $\alpha_P(T)$, and $\phi_J(T)$.  Each of these TPCs is a four-parameter model (for example, taking: optimal temperature $T_o$, maximum performance $P_{max}$, breadth up to $a_+$ and beyond $a_-$ optimum temperature.  

A sensible way to reduce this will be to assume some consistency in TPC across life-history stages.  For example, we might assume that the optimal temperature is likely to be the same for all three of these curves.  We might also assume that the breadth of the curves for transition probabilities are the same and that really the only thing that changes across the functions for transition probabilities $\alpha$, is the maximum performance.

Thus, we might look to fit a model with some of the temperature-dependence parameters shared across these temperature dependent vital rates.  

To fit these models, of course, we need data. What can we extract from the literature?

#### $\alpha_J$ data and priors

For $\alpha_J$ we have the data from @umeda2019 (plotted above).

```{r umeda_alpha_j_data}
alpha_j_temp_umeda <- c(18, 20, 25, 30, 32)
alpha_j_rate_umeda <- c(0.013, 0.020, 0.0258, 0.0421, 0.0206)
```

We might also make use of the data from @walgama2007.  These authors measured developmental rates for egg, larva and pupa separately (which together gives $\phi_j$).  Their data (eggs and larvae reported in their tables 1, 2, and larvae rates extracted from Fig 1.) are as follows:

```{r walgama_alpha_j_data}
alpha_j_temp_walgama <- c(15, 18, 20, 22, 25, 28, 30, 32)
egg_dev_time_walgama <- c(Inf, 23.5, 18.5, 13, 7.3, 5.5, 5, 10.9)
larva_dev_time_walgama <- c(NA, 1/0.0288, 1/0.0399, NA, 1/0.0688, 1/0.140, NA, NA) # extracted from figure
pupae_dev_time_walgama <- c(Inf, 15, 14, 9, 7.5, 6, 4, 7)
total_juv_dev_time_walgama <- egg_dev_time_walgama + larva_dev_time_walgama + pupae_dev_time_walgama
alpha_j_rate_walgama <- 1/total_juv_dev_time_walgama
alpha_j_rate_walgama[1] <- 0

# plot the two datasets against each other
plot(alpha_j_rate_walgama~alpha_j_temp_walgama)
points(alpha_j_rate_umeda~alpha_j_temp_umeda, pch=17)
```

These look to be astonishingly consistent, really.  We will combine these two sets to estimate our TPC for alpha_j.

```{r alpha_J_TPC}
temp <- c(alpha_j_temp_umeda, alpha_j_temp_walgama)
rate <- c(alpha_j_rate_umeda, alpha_j_rate_walgama)
TPCmatrix <- na.omit(cbind(temp, rate))
alpha_J_fit <- TPC.q.fit(TPCmatrix, in.acc = 8, hessian = TRUE)
pars <- alpha_J_fit$par

plotTemp <- seq(12, 35, 0.2)
survProbPred <- TPC.q(plotTemp, rmax = pars[1], Trmax = pars[2], acc = pars[3], dec.prop = pars[4])

plot(rate~temp)
lines(survProbPred~plotTemp)
```

Which looks decent.  Now we just need some errors on the parameter estimates, which we should be able to get from the inverse of the Hessian.

```{r alpha_J_TPC_priors}
par_names_TPC <- c(quote(P[max]), quote(T[0]), quote(a[plus]), quote(a[minus]))
alpha_J_pars_sd <- sqrt(diag(solve(alpha_J_fit$hessian)))
alpha_J_priors <- cbind(pars, alpha_J_pars_sd)
rownames(alpha_J_priors) <- par_names_TPC
alpha_J_priors
```

#### $\alpha_P$ data and priors

There is a lack of real data for this prior.  @walgama2007 make their estimate assuming a pre-oviposition period of 8 days at optimal temperature, and a minimum developmental temperature of 15 degrees.  If we take the above curve as a template, we would switch out $P_{max}$ with `r 1/8`.  For the sake of making a prior, let us assume that our uncertainty scales with the mean estimate.  This would yield a prior for the $P_{max, \alpha_P}$ as follows.

```{r alpha_P_TPC_priors}
alpha_P_priors <- alpha_J_priors
colnames(alpha_P_priors) <- c("pars", "alpha_P_pars_sd")
alpha_P_priors["P[max]", 1] <- 1/8
alpha_P_priors["P[max]", 2] <- sqrt(1/8/alpha_J_priors["P[max]", 1]*(alpha_J_priors["P[max]", 2]^2))
alpha_P_priors
```

#### $\phi_J$ data and priors

The only data here come from @walgama2007 and are given above (held in `sDat`).  Let's repeat that analysis, but calculate the Hessian along the way.

```{r phi_j_TPC_priors}
phi_J_fit <- TPC.q.fit(TPCmatrix, in.acc = 0.3, hessian = TRUE)
phi_J_pars <- phi_J_fit$par
phi_J_pars_sd <- sqrt(diag(solve(phi_J_fit$hessian)))
phi_J_priors <- cbind(phi_J_pars, phi_J_pars_sd)
rownames(phi_J_priors) <- par_names_TPC
phi_J_priors
```
 
#### $\phi_P$, $\phi_A$ data and priors
 
Pre-adult and adult survival appears to be less affected by temperature.  Let us assume that these are equal and constant over time.  @gadd1941 estimates a maximum laying lifespan for a female as 49 days, with a rough estimate of the mean laying lifespan of 32 days. This would yield a per day mortality rate of 1/32.  This equates to a per day survival probability of $\phi_P = \phi_A=$ `r exp(-1/32)`  We have no real way to put bounds on this, so will pick a beta distribution with this mean and 95% bounds approximately $\pm 0.3$.

We can assume pre-adult survival is the same as adult survival. This leaves us only with the survival associated with dispersal, $\phi_\mu$, which we have no measure for, we will make this a minimally informative beta distribution.

## References
