---
title: "PSHB Models"
output: html_document
date: '`r format(Sys.Date(), "%d/%m/%Y")`'
bibliography: 'PSHBRefs.bib'
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo=TRUE)
options(scipen = 10, digits = 2)
```

## Basic life history

Polyphagous shothole borer, *Euwallacea fornicatus*, is a pest of a wide range of plants and has recently expanded its range from south-east Asian to Africa, the Americas, and Australia.

Mated females alight on a host plant and bore a small gallery in the hodt tree tissue. They seed an ambrosia fungus into this gallery and spend some amount of time building the gallery before laying eggs. After hatching juveniles, moult through several larval instars, pupate and eventually eclose as adults. The sex ratio of adults is m:f 1:3-5. Males are wingless and never leave the gallery. Females are winged. They mate with their brothers, and then exit the gallery to found a new gallery (on the same or different host). The rate of egg laying declines steadily over time, with females eventually dying in the gallery rather than forming a new one.

Development of all stages is strongly temperature dependent. There is some evidence of developmental deaths occurring between hatching and pupation. Other sources of mortality are likely at the dispersal phase (predation, misadventure), otherwise the life history is fairly safe from predation.

## Our treatment

### Discrete time model

The life history suggests we might profitably partition the population of females as:

  * $J(t,x)$, the juvenile stages (eggs through pupae): some mortality in the nest, no dispersal 
  * $P(t,x)$, the pre-adult stage (the dispersal phase): dispersal, with some mortality 
  * $A(t,x)$, the adult stage (breeding females): no dispersal, mortality through senescence.

We measure time, $t$, in days. We also partition our population among host trees, $x$.

#### Within host dynamics

Within a single host (i.e.ignoring $x$), and using the same state variables above we would have:


\begin{align}
J(t+1) &= \phi_J(1-\alpha_J)J(t) + fA(t\\
P(t+1) &= \phi_J\alpha_JJ(t) + \phi_P(1-\alpha_P)(1-\mu)P(t) + 0 \\
A(t+1) &= 0 + \phi_P\alpha_P(1-\mu)P(t) + \phi_AA(t)
\end{align}

Where:

  * $f(t)$ is the mean fecundity of a female in the population in an interval of time 
  * $\alpha_i(t)$ are transition probabilities, describing transition between states
  * $\phi_i(t)$ are survival probabilities, describing probability of surviving an interval of time in each state 
  * $\mu(t)$ is the probability of dispersing away from the host tree.

Although we drop the notation, all these are functions of time, to capture the effect of temperature on the vital rates.  In matrix form, we first define the column vector $\mathbf{n}= \{J, P, A\}$. Then we have:

$$
\begin{gather}
 \mathbf{n}(t+1) =
  \begin{bmatrix}
   \phi_J(1-\alpha_J) & 0 & f \\
   \phi_J\alpha_J & \phi_P(1-\alpha_P)(1-\mu) & 0 \\
   0 & \phi_P\alpha_P(1-\mu) & \phi_A
  \end{bmatrix}
 \mathbf{n}(t)
\end{gather}
$$

The matrix shown is the within-population transition matrix, $\mathbf{W}$.

Note also that in the within-host model, we currently neglect to account for the arrival of pre-adults from other hosts. We will come back to this in due course.

Note that there is no density dependence in the model for now. The obvious place to put this would be on $\phi_J$. That is, we could make $\phi_J$ some negative function of total number of juveniles produced in the host. Something like,

$$
\phi_J(t, x) = e^{-b(x)\sum_t{fA(t)}}.
$$

Which allows each host to have its own 'carrying capacity'. Perhaps simpler would be a threshold such that,

$$
\begin{gather}
  \phi_J(t, x) = 
  \begin{cases}
    \phi_J & \text{if } \sum_t{fA(t, x)} < C(x) \\
    0 & \text{if } \sum_t{fA(t, x)} > C(x)
  \end{cases}
\end{gather}
$$ 

Here, each host could have its own threshold, $C(x)$, for production of beetles. Beyond this threshold, the host resource is used up and $\phi_J$ collapses to 0.

#### Between host dynamics

We now seek to extend the model to space, to capture dynamics across multiple hosts. To do this, we need to describe dispersal dynamics. We assume that $\mu P(t, x)$ leave each host at each time period. We might further assume that there is some survival rate associated with dispersing between hosts and establishing a new gallery. Let's call that $\phi_\mu$. Thus, the number of effective dispersers from a given host at a given time is: $\phi_\mu\mu P(t, x)$.

We will make this model spatially explicit. So the probability of arriving at a given host will depend on the distance between source and recipient host, let us call that distance, $d_{xy}$, the distance between host $x$ and host $y$. We can imagine a pairwise distance matrix, $\mathbf{D}$, which is an $h\times h$ matrix, where $h$ is the total number of hosts. The probability of moving to a given host, is given by a 2-dimensional dispersal kernel, $k(d_{ij})$. There are a large number of possible functions we could use for $k()$; let us leave it general for now.

It is tempting at this stage to specify dispersal as a stochastic process. The reason for this is that without stochasticity we can have fractions of individuals arriving at sites and growing; we would need to control this in some way. With stochasticity, we can arrange things to only have integer values of immigrants. For example, we might examine the fate of migrants from host $y$. These will disperse to other locations, $x\neq y$ in proportion to the distances between sites as defined by the kernel, $k(d_y)$ where $d_y$ is the vector of distances between $y$ and all other sites. We can draw the realised number of dispersers arriving to each site, $x$ as,

$$
\mathbf{m_y} \sim \text{Multinom}(m_y, k(d_y))
$$ where $m_y$ is the realised number of migrants emerging from site $y$:

$$
m_y \sim \text{Binom}(P(t, y), \phi_\mu\mu).
$$

We can get away with this without further adjustments to the within-host matrix, because we assume that it is survival during dispersal that introduces the stochastic element here.

By stepping through all sites, $y$, we arrive at a realised dispersal matrix, $\mathbf{M}$. And the realised number of immigrants arriving at site $x$ is then the vector given by the row sums of $\mathbf{M}$: $\mathbf{m_+} = \sum_y \mathbf m_y$.

### Within and between host model

So, now we need to bring this all together. First step is to re-define the poopulation vector as $\mathbf{n} = \{\mathbf{n_1}, ..., \mathbf{n_h} \} = \{J_1, P_1, A_1, J_2, P_2, A_2, ..., J_h, P_h, A_h,\}$, thus now containing all individuals at all stages across all the hosts in the population.

We then define the $h \times h$ diagonal block matrix, $\mathbf{T}$:

$$
\mathbf{T} = 
 \begin{bmatrix}
  \mathbf{W} & 0  & 0 \\
  0 & \ddots & 0 \\
  0 & 0 & \mathbf{W}
 \end{bmatrix}.
$$ 

Here each $0$ denotes a $3\times 3$ matrix of zeroes, and $\mathbf{W}$ is the within-host transition matrix.

Normally, we would place dispersal into the off-diagonal elements of this matrix, but we have made dispersal stochastic, so easier to leave it out. Instead, we then need to get $\mathbf m_+$ into the same length as our redefined $\mathbf{n}$. So we define a new vector,

$$
\mathbf{m} = \{0, 0, \mathbf{m_{+1}}, 0, 0, \mathbf{m_{+2}}, ..., 0, 0, \mathbf{m_{+h}} \}
$$ Our final model then becomes:

$$
\begin{gather}
 \mathbf{n}(t+1) = \mathbf{T}\mathbf{n}(t) + \mathbf{m}(t).
\end{gather}
$$
## Estimating life-history parameters

### Transition probabilities

```{r parameterSet, echo=FALSE}
delta_t <- 20
wal_juv <- 266
wal_pre <- 136
```

### Transition probabilities

Several sources have estimated the 'thermal constant', $K$ for PSHB and related species.  $K$ is the cumulative number of degree days above some minimum temperature that it takes for development to complete.  It is a standard approach in entomology. So, if the minimum temperature is 15 degrees and we have a day where the mean temperature is 20 degrees, then we have accumulated 5 degree days towards development.  This is what is meant by $K$.

The results for $K$ in *Euwallacea* are nicely summarised in @dodge2019.

| Species | $T_{min}$ | $T_{opt}$ | $T_{max}$ | $K$ | Reference |
|:--:|:--:|:--:|:--:|:--:|:--:|
| TSHB | 15 | 30 | 32 | 373 | @walgama2007 |
| PSHB | 13.34 | 27.51 | 33.08 | 398 | @umeda2019 |
| KSHB | 12.77 | 28.04 | 31.99 | 318 | @dodge2019 |

Note Dodge's estimate is for (nearly) the full life cycle from egg to emergent adult (*i.e.*, $\alpha_J\alpha_P$). Umeda's estimate is for the transition from juvenile to pre-adult (*i.e.*, $\alpha_J$).  Walgama's estimate is for the full life-cycle.  Dodge and Walgama's studies are best, and Walgama's gives the clearest breakdown of rates.

This degree.day format is not actually useful to us, because while development can be seen as constant with respect to degree days, survival has only ever been estimated as a function of temperature.  My initial thought was to change the time axis of our model to degree.days and use the resulting constant transition probabilities.  This would have been computationally most efficient. Problem with this is that the degree.day axis obscures absolute temperature, and survival has only ever been estimated with regard to absolute temperature.  So if we want a model with both transition and survival probabilities, we need to stick to the absolute time axis.  

Thus, we need to step back from the degree day results, and look for the per day rates that went into the estimation of $K$.  @umeda2019 provide the following data in their paper.

```{r umedaData, fig.cap="Data on the rate of development from egg to pre-adult, from @umeda2019. Line is fitted ignoring the result for 32 degrees, which is considered above the thermal threshold."}
temp <- c(18, 20, 25, 30, 32)
rate <- c(0.013, 0.020, 0.0258, 0.0421, 0.0206)
plot(rate~temp, xlab = "Temperature", ylab = "Development rate (per day)")
ss <- 1:4
mod <- lm(rate[ss]~temp[ss])
abline(reg=mod)
```

We can use this to build a function returning $\alpha_J$. First, the estimated coefficients of the linear model:


```{r}
coef(mod)
```

Then we can build a function.

```{r alpha_J_temp}
alpha_J_temp <- function(temperature, lower = 15, upper = 31){
  rate <- -0.0273 + 0.0023*temperature
  prob <- 1-exp(-rate)
  ifelse(temperature>lower & temperature < upper, prob, 0)
}
```

Let's see what $\alpha_J$ looks like across a range of temperatures.

```{r alpha_J_plot}
temps <- seq(10, 35, 0.1)
alpha_J <- alpha_J_temp(temps)
plot(alpha_J~temps, 
     type = "l", 
     xlab = "Temperature", 
     ylab = quote(italic(alpha[J])),
     bty = "l")
```

@umeda2019 do not provide estimates of transition rates through the pre-adult stage.  Instead, here we turn to @walgama2007, who estimate 136 degree days with a minimum of 15 degrees for this stage.  These values can be mapped back to a linear function of development rate against temperature, as follows.

```{r alpha_P_estimate}
alpha_P_temp <- function(temperature, lower = 15, upper = 31){
  DD <- 136 #degree days from walgama
  beta <- 1/DD # implied slope of rate on temperature
  alpha <- -15*beta # implied intercept
  rate <- alpha + beta*temperature
  prob <- 1-exp(-rate)
  ifelse(temperature>lower & temperature < upper, prob, 0)
}

alpha_P <- alpha_P_temp(temps)
plot(alpha_P~temps, 
     type = "l", 
     xlab = "Temperature", 
     ylab = quote(italic(alpha[P])),
     bty = "l")

```


### Survival probabilities

Survival probability is also clearly temperature dependent.  

## References
