---
title: "Supplementary Material"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo=TRUE)
options(scipen = 10, digits = 2)
```

## PSHB Thermal Performance Curves (TPCs)

Three of our PSHB life-history parameters are temperature-dependent: the daily probabilities of juvenile transition ($\alpha_J$), pre-oviposition transition ($\alpha_P$) and juvenile survival ($\phi_J$). To generate functions for temperature-dependent transition and survival, we fit temperature performance curves (TPCs) to empirical PSHB data. TPCs describe the relationship between temperature, $T$, and performance, $P$. In this instance, performance is the probability of transition or survival. As described in Phillips *et al.*, (2014), a TPC function is given as a meeting of two Gaussian functions:

$$
\begin{gather}
  P(T) = 
  \begin{cases}
    \ P_{max}e^{- \frac{-(T-T_{opt})^2} {2c^2} } & \text{ for } T<T_{opt}\\
    P_{max}e^{- \frac{-(T-T_{opt})^2} {2(cd)^2} } & \text{ for } T \ge T_{opt}
  \end{cases}
\end{gather}\text{(Supp. Eq. 1),}
$$ 
where $P_{max}$ is the maximum performance value and $T_{opt}$ is the optimum temperature (*i.e.*, the temperature at $P_{max}$). Performance increases with temperature up to $T_{opt}$, according to $c$. At temperatures above $T_{opt}$, performance decreases, as given by $d$.

TPCs for $\alpha_J$, $\alpha_P$ and $\phi_J$ were generated, and parameter values for Supp. Eq. 1 were estimated  (described below). When running the population model, the daily tree temperature was input as $T$ in Equation 1, to calculate the values of $\alpha_J$, $\alpha_P$ and $\phi_J$ to use for that time-step.


### $\alpha_J$ TPC

The daily probability of juvenile PSHB transitioning to the preoviposition stage is given by $\alpha_J$. We estimated $\alpha_J$ using data from Umeda and Paine, (2019), who monitored PSHB development (from egg to the pre-oviposition stage) under temperatures ranging 18 – 32 °C. Development rates are summarised in Supplementary Table 1. 

We also used data from Walgama and Zalucki, (2007), who recorded development times for the tea shot hole borer (TSHB, within the *E. wallacea* species complex) at 15 – 32° C. Development times for eggs, larvae and pupae were obtained from Table 1, Figure 1, and Table 2, respectively (Supp. Table 1). For each temperature, we summed development times to calculate the total duration of the juvenile stage, $\delta$, and converted this to a daily transition rate: $\alpha_{J} = \delta^{-1}$ (Supp. Table 1). Transition rates from the two studies are plotted in Supp. Figure 1.


#### Supplementary Table 1
###### The development times of PSHB eggs, larvae and pupae at different temperatures (from Walgama & Zalucki, 2007), used to estimate daily juvenile transition rates; and daily juvenile transition rates at different temperatures given in Umeda & Paine, (2019).   

| Temperature (°C)| Egg development time (days) | Larva development time (days) | Pupa development time (days) | Juvenile transition rate (Walgama & Zalucki) | Juvenile transition rate (Umeda & Paine) |
|:--:|:--:|:--:|:--:|:--:|:--:|
| 15 | no hatching |  | no emergence | 0 |  |
| 18 | 23.5 | 34.72 | 15 | 0.014 | 0.013 |
| 20 | 18.5 | 25.06 | 14 | 0.017 | 0.02 |
| 22 | 13 |  | 9 |  |  |
| 25 | 7.3 | 14.53 | 7.5 | 0.034 | 0.0258 |
| 28 | 5.5 | 7.14 | 6 | 0.054 |  |
| 30 | 5 |  | 4 |  | 0.0421 |
| 32 | 10.9 |  | 7 |  | 0.0206 |



We fit a TPC to the empirical data using... 
*how to describe these functions??*

```{r TPC.q.fit}
# Equation 1:
TPC.q<-function(Tb, rmax=10, Trmax=28, acc=9, dec.prop=0.5){
	lhs<-rmax*exp(-(Tb-Trmax)^2/(2*acc^2))
	rhs<-rmax*exp(-(Tb-Trmax)^2/(2*(acc*dec.prop)^2))
	ifelse(Tb<Trmax, lhs, rhs )
}

TPC.q.LL<-function(pars, t.data, w.data){
  E.w<-TPC.q(Tb=t.data, rmax=pars[1], Trmax=pars[2], acc=pars[3], dec.prop=pars[4])
  D.w<-w.data-E.w
  D.w<-dnorm(D.w, mean=0, sd=sd(D.w), log=TRUE) 
  -sum(D.w)
}

TPC.q.fit<-function(Tw.matrix, in.acc=10, in.dec.prop=0.5, ...){
	in.rmax=max(Tw.matrix[,2])
	in.Trmax=Tw.matrix[Tw.matrix[,2]==in.rmax,1]
	fit<-optim(fn=TPC.q.LL, par=c(rmax=in.rmax, Trmax=in.Trmax, acc=in.acc, dec.prop=in.dec.prop),
		t.data=Tw.matrix[,1], w.data=Tw.matrix[,2], method="L-BFGS-B",
		lower=c(0, 10, 1, 0.001), upper=c(1.5*in.rmax, 45, 80, 1), ...)
	fit
}
```



```{r alpha_j_data_TPC, echo=FALSE}
alpha_j_temp_umeda <- c(18, 20, 25, 30, 32)
alpha_j_rate_umeda <- c(0.013, 0.020, 0.0258, 0.0421, 0.0206)

alpha_j_temp_walgama <- c(15, 18, 20, 22, 25, 28, 30, 32)
egg_dev_time_walgama <- c(Inf, 23.5, 18.5, 13, 7.3, 5.5, 5, 10.9)
larva_dev_time_walgama <- c(NA, 1/0.0288, 1/0.0399, NA, 1/0.0688, 1/0.140, NA, NA) # extracted from figure
pupae_dev_time_walgama <- c(Inf, 15, 14, 9, 7.5, 6, 4, 7)
total_juv_dev_time_walgama <- egg_dev_time_walgama + larva_dev_time_walgama + pupae_dev_time_walgama
alpha_j_rate_walgama <- 1/total_juv_dev_time_walgama
alpha_j_rate_walgama[1] <- 0
temp <- c(alpha_j_temp_umeda, alpha_j_temp_walgama)
rate <- c(alpha_j_rate_umeda, alpha_j_rate_walgama)

source("../src/TPCFunctions.R")

TPCmatrix <- na.omit(cbind(temp, rate))
alpha_J_fit <- TPC.q.fit(TPCmatrix, in.acc = 8, hessian = TRUE)
pars <- alpha_J_fit$par

plotTemp <- seq(12, 35, 0.2)
survProbPred <- TPC.q(plotTemp, rmax = pars[1], Trmax = pars[2], acc = pars[3], dec.prop = pars[4])

plot(alpha_j_rate_walgama~alpha_j_temp_walgama,
     xlab = expression(paste("Tree temperature (",degree,"C)")), 
     ylab = expression('Juvenile transition probability ( '*alpha[j]*' )'))
points(alpha_j_rate_umeda~alpha_j_temp_umeda, pch=17)
lines(survProbPred~plotTemp)
```


#### Supplementary Figure 1
###### Juvenile transition probabilities at different temperatures, given by Umeda & Paine (triangles) and Walgama & Zalucki (circles), with TPC fitted to the data.




The parameter estimates for the TPC ($P_{max}$, $T_{opt}$, $c$, $d$) and their standard deviations were:

```{r alpha_J_TPC_priors, echo=FALSE}
par_names_TPC <- c(quote(P[max]), quote(T[opt]), quote(c), quote(d))
SD <- sqrt(diag(solve(alpha_J_fit$hessian)))
alpha_J_priors <- cbind(pars, SD)
rownames(alpha_J_priors) <- par_names_TPC
colnames(alpha_J_priors) <- c('Estimate', 'SD')
alpha_J_priors
```
*How to describe how we get these??*


### $\alpha_P$ TPC

The transition probability for the pre-ovipositing stage is given by $\alpha_P$. There is limited empirical data for this stage. In Walgama and Zalucki (2007),  the pre-oviposition period for TSHB was estimated to be 136 degree-days, or 8 days at $T_{opt}$, and with a minimum temperature of 15°C. We therefore assigned $P_{max}=1/8$. We assumed that the other TPC parameters ($T_{opt}$, $c$, $d$) were the same as for  


### $\phi_J$ TPC

The daily survival probability of juveniles, $\phi_J$, is also likely temperature-dependent. Walgama and Zalucki (2007) report the total mortality of eggs and pupae at different temperatures. We extracted these data (from Figure 2 of that study), and provide it in Supp. Table 2.

We used these values of total proportion mortality, along with the total duration of each stage (given Supp. Table 1), to calculate the daily probability of survival:

\begin{align}
\phi = e^{-(\frac{-log(1-\lambda)}{\delta})}
\end{align}

where $\lambda$ is the total mortality of the stage, and $\delta$ is the total duration of the stage. Estimates of $\phi_J$ from the empirical data are given in Supp. Table 2, and plotted in Supp. Figure 2.

```{r phi_J_data}
# Don't need this, just showing the working of phi calculation

library(readr)
library(dplyr)
library(lubridate)
sDat <- read.csv(file = "../dat/walgamaFig2DataExtract.csv")
sDat$Mortality <- sDat$Mortality/100 # convert to probability of mortality over n days
sDat$Mortality[sDat$Mortality>1] <- 1 # catch the 1.002s
mortRate <- -log(1-sDat$Mortality)/sDat$Days # convert to a rate
mortProb <- 1-exp(-mortRate) # convert to a daily probability
```

#### Supplementary Table 2
###### Total proportion mortality of eggs and pupae at different temperatures (from Walgama & Zalucki 2007), and the daily survival probabilities estimated from these data.

| Temperature (°C)|  Total egg mortality | Total pupa mortality | Daily egg survival | Daily pupa survival |
|:--:|:--:|:--:|:--:|:--:|
| 15 |  1 | 1 |  |  | 
| 18 | 0.2 |0.278 | 0.991 |  0.979 | 
| 20 | 0.163 |0.147 | 0.990 |  0.989 | 
| 22 |  0.14 |0.124 | 0.988 |   0.985 |
| 25 | 0.12 |0.051 | 0.983 |  0.993 | 
| 28 |  0.021 |0.002 | 0.996 |  0.999 | 
| 30 | 0.051 |0.014 | 0.990 |   0.996 | 
| 32 | 0.421 |0.041  | 0.951 |  0.994 | 


As above, we fitted a TPC to the empirical data (Supp. Fig. 2) and parameter values for the TPC function.

```{r plot_phi_J_TPC, echo=FALSE}
survProb <- 1-mortProb
sDat <- cbind(sDat, mortProb, survProb)

TPCmatrix <- cbind(sDat$Temperature, survProb)
TPCmatrix <- na.omit(TPCmatrix)
pars <- TPC.q.fit(TPCmatrix, in.acc = 0.3)$par
temps <- seq(10, 35, 0.1)
survProbPred <- TPC.q(temps, rmax = pars[1], Trmax = pars[2], acc = pars[3], dec.prop = pars[4])
plot(survProb~sDat$Temperature,
     xlab = expression(paste("Temperature (",degree,"C)")), 
     ylab = expression('Juvenile survival probability ( '*phi[J]*' )'))
lines(survProbPred~temps)
```
#### Supplementary Figure 2
###### Juvenile survival probabilities at different temperatures given by Walgama & Zalucki, with TPC fitted to the data.

```{r phi_j_TPC_priors_2, echo=FALSE}
phi_J_fit <- TPC.q.fit(TPCmatrix, in.acc = 0.3, hessian = TRUE)
phi_J_pars <- phi_J_fit$par
phi_J_pars_sd <- sqrt(diag(solve(phi_J_fit$hessian)))
phi_J_priors <- cbind(phi_J_pars, phi_J_pars_sd)
rownames(phi_J_priors) <- par_names_TPC
colnames(phi_J_priors) <- c("Estimate", "SD")
phi_J_priors
```


## Tree temperature predictions

Tree temperature, $T_{tree}$, was a function of the maximum air temperature, $T_{max}$, and the soil temperature, $T_{soil}$:

\begin{align}
T_{tree} = p_tT_{max} + (1-p_t)T_{soil} &\text{,}
\end{align}

where $p_t$ can take a value between 0 and 1, and is a function of the relative humidity, $H$. Specifically, $p_t$ was taken as the *inverse logit(?)* of a linear function:

\begin{align}
p_t = \text{inverse logit}(bH+c)
\end{align}

In the population model:
``` {r temp_fun_model}
#  logit.p <- int + beta*rh_tmax # int & beta are hard-coded, rh_tmax is gotten through SILO
#  p <- plogis(logit.p)
#  mean_temp <- p*air_tmax + (1-p)*ma30 # mean_temp then used to estimate transition & survival probs 
```


To estimate values for $b$ and $c,$ we fitted this model in a Bayesian framework using HMC, implemented in the greta package (Golding, 2019).


``` {r get_temp_data, echo=FALSE}
source("../src/temperatures/get-calibration-data_markdown.R")
```

``` {r tree_temp_greta}
# Where merge_temp is data set combining sap flow and SILO data
merge_temp <- na.omit(merge_temp) # remove NAs (otherwise greta doesn't work?)

treeTemps <- as_data(merge_temp$mean_d) # mean daily probe temp
rh <- as_data(merge_temp$rh_tmax) # relative humidity
air <- as_data(merge_temp$air_tmax) # max air temp
soil <- as_data(merge_temp$soil) # soil temp (30-day-average of air temp)

# describe predictor function
# priors
int <- normal(0, 3)
beta <- normal(0, 3)
sd <- lognormal(0, 3)

p <- ilogit(int + beta*rh) # rh predicts p

mean_temp <- p*air + (1-p)*soil # predicted temp is weighted average of air/soil according to p

distribution(treeTemps) <- normal(mean_temp, sd)
  
gMod <- model(int, beta, sd)

plot(gMod) # Plot representation of model

draws <- mcmc(gMod, n_samples = 1000, chains = 4)
bayesplot::mcmc_trace(draws)

# make a plot of prediction bounds against data
draws_plot <- calculate(mean_temp, 
                        nsim = 1000,
                        values = draws)
pred_temp <- apply(draws_plot[[1]], 2, mean)
ci_temp <- apply(draws_plot[[1]], 2, quantile, prob = c(0.025, 0.975))
plot_data <- cbind(DOY = merge_temp$DOY, pred_temp, t(ci_temp))
plot_data <- plot_data[order(plot_data[,"DOY"]),]
```

*How much of this to include??*

``` {r tree_temp_plot}
plot(treeTemps~DOY, data = merge_temp,
     ylab = expression(paste("Tree temperature (",degree,"C)")), 
     xlab = "Day of year")
lines(plot_data[,"2.5%"]~plot_data[,"DOY"], col = "red")
lines(plot_data[,"97.5%"]~plot_data[,"DOY"], col = "red")
```
#### Supp. Figure
###### Mean daily tree temperatures per day of year, as recorded by sap flow loggers (points), and as predicted by our Bayesian model (95% upper and lower CI, red lines).

The mean estimates for $b$ and $c$ in the humidity function were calculated:
``` {r pt_coeffs}
pt_coef <- summary(draws)$statistics
rownames(pt_coef) <- c("c", "b", "SD")
pt_coef
```

and then used in the population model to calculate $p_t$ at each time-step, according to humidity.

